Scaffolder - Software for \change{Reproducible}{Manual} Genome Scaffolding.

### 2011-08-13 src/article.tex
Change manuacript title
Text fixes

\title{Scaffolder - Software for Reproducible Genome Scaffolding.} %{{{1

@@

Assembly software takes the nucleotide reads produced by sequencing hardware
and, in the ideal case, outputs a single complete genome sequence composed of
these individual fragments. An analogy for this process is a jigsaw puzzle:
each nucleotide read represents a single piece, and the final genome sequence
is the completed puzzle. Sequences of repetitive nucleotide `repeat' regions
within the genome, or biased and incomplete sequencing data, may prevent the
genome being assembled into a continuous sequence. This may be due to not
enough, or multiple different overlaps between reads and is analogous to
missing pieces in the jigsaw or pieces that fit to multiple other pieces. \pb

The advent of high throughput sequencing methods has led to a renewed interest
in algorithms to solve the problem of genome assembly \cite{pop2008,pop2009}.
Nonetheless, due to the complexity of merging large numbers of overlapping
reads genome assembly software may be unable to produce a complete sequence.
Instead, the algorithm may generate several large assembled regions of
sequence (`contigs') composed of the many individual reads that represent
assembly into the longest possible sequences, given the data. Nonetheless,
these contigs represent a fragmented picture of the genome and require
additional work to join together into a complete sequence. \pb

The process of ameliorating a draft sequence into a finished continuous
sequence can be expensive in terms of time and laboratory effort, although the
genomic data present in a set of generated contigs may be sufficient for many
research questions \cite{branscomb2002}. Nevertheless, a continuous
high-quality `finished' genome sequence does provide a greater depth of
information, such as complete resolution of repeat regions and precise
estimates of distances between genomic elements \cite{parkhill2002,fraser2002}.
The process of joining these contigs together to form a continuous genome
sequence is called the `scaffolding' or `finishing' stage and is the focus of
the software described in this article. \pb

@@

The sequencing process generates reads from either strand of the DNA helix and
the resulting contigs constructed from these reads may represent either the
5'~$\rightarrow$~3' or 3'~$\rightarrow$~5' DNA strands with respect to the
origin of replication. Orientating all contigs to point in the same direction
requires reverse complementing sequences where necessary. In the case of
microbial genomes this orientation will be to the 5'~$\rightarrow$~3'
direction following the direction of genome replication.

@@

The process of resolving a genome sequence and building a scaffold uses wet
laboratory methods, \emph{in silico} methods, or a combination of both. An
example of a computational method might use the existing paired-read data from
the sequencing stage where reads were produced in pairs at a known distance
apart in the genome. The occurrence of paired reads in separate contigs can be
used to probabilistically estimate the order and distance between these
contigs. Alternatively, laboratory methods may use PCR to amplify the unknown
DNA in a gap region, then use traditional Sanger sequencing to determine the
sequence of this gap. Computational methods are more preferable as they are
less costly in laboratory time and materials compared to manual gap resolution,
and use the available data that exists following sequencing
\cite{nagarajan2010}. Finally when the scaffold cannot be completely resolved,
\emph{in silico} software packages exist to suggest the likely primers
necessary for PCR amplifying the sequence in gap regions \cite{gordon2001}. \pb

@@

Repeat regions may also be responsible for multiple gaps when building
a genome sequence; tandemly repeated nucleotide regions in the genome produce
multiple reads with similar sequence. As many assembly algorithms rely on
sequence overlaps between reads to build a contig, the similarity between
repeat-region reads can result in the assembly collapsing into an artificially
short sequence or being ignored by more conservative assembly algorithms. Such
regions can be resolved by using algorithms specifically reassemble the
collapsed repeat region correctly \cite{mulyukov2002,koren2010}, while
a related approach uses unassembled sequence reads matching the regions around
a scaffold gap to construct a uniquely overlapping set of reads across it
\cite{tsai2010}. \pb

>>>>

\title{Scaffolder - Software for \change{Reproducible}{Manual} Genome
Scaffolding.}
%{{{1

@@

Assembly software takes the nucleotide reads produced by sequencing hardware
and, in the ideal case, outputs a single complete genome sequence composed of
these individual fragments. An analogy for this process is a jigsaw puzzle:
each nucleotide read represents a single piece, and the final genome sequence
is the completed puzzle. Sequences of repetitive nucleotide `repeat' regions
within the genome, or biased and incomplete sequencing data, may prevent the
genome being assembled into a continuous sequence. This may be due to
\change{not enough}{insufficient}, or multiple different overlaps between reads
and is analogous to missing pieces in the jigsaw or pieces that fit to multiple
other pieces. \pb

The advent of high\add{-}throughput sequencing methods has led to a renewed
interest in algorithms to solve the problem of genome assembly
\cite{pop2008,pop2009}. Nonetheless, due to the complexity of merging large
numbers of overlapping reads genome assembly software may be unable to produce
a complete sequence. Instead, the algorithm may generate several large
assembled regions of sequence (`contigs') composed of the many individual reads
that represent assembly into the longest possible sequences, given the data.
Nonetheless, these contigs represent a fragmented picture of the genome and
require additional work to join together into a complete sequence. \pb

The process of \change{ameliorating a draft sequence into a finished continuous
sequence}{finishing a genome sequence} can be expensive in terms of time and
laboratory effort, although the genomic data present in a set of generated
contigs may be sufficient for many research questions \cite{branscomb2002}.
Nevertheless, a continuous high-quality `finished' genome sequence does provide
a greater depth of information, such as complete resolution of repeat regions
and precise estimates of distances between genomic elements
\cite{parkhill2002,fraser2002}. The process of joining these contigs together
to form a continuous genome sequence is called the `scaffolding' or `finishing'
stage and is the focus of the software described in this article. \pb

@@

The sequencing process generates reads from either strand of the DNA helix and
the resulting contigs constructed from these reads may represent either the
5'~$\rightarrow$~3' or 3'~$\rightarrow$~5' DNA strands with respect to the
origin of replication. Orientating all contigs to point in the same direction
requires reverse complementing sequences where necessary. In the case of
\change{microbial}{archaeal and bacterial} genomes this orientation will be to
the 5'~$\rightarrow$~3' direction following the direction of genome
replication.

@@

The process of resolving a genome sequence and building a scaffold uses wet
laboratory methods, \emph{in silico} methods, or a combination of both. An
example of a computational method might use the existing paired-read data from
the sequencing stage where reads were produced in pairs at a known distance
apart in the genome. The occurrence of paired reads in separate contigs can be
used to \change{probabilistically estimate}{estimate probabilistically} the
order and distance between these contigs. Alternatively, laboratory methods may
use PCR to amplify the unknown DNA in a gap region, then use traditional Sanger
sequencing to determine the sequence of this gap. Computational methods are
\remove{more} preferable as they are less costly in laboratory time and
materials compared to manual gap resolution, and use the available data that
exists following sequencing \cite{nagarajan2010}. Finally when the scaffold
cannot be completely resolved, \emph{in silico} software packages exist to
suggest the likely primers necessary for PCR amplifying the sequence in gap
regions \cite{gordon2001}. \pb

@@

Repeat regions may also be responsible for multiple gaps when building
a genome sequence; tandemly repeated nucleotide regions in the genome produce
multiple reads with similar sequence. As many assembly algorithms rely on
sequence overlaps between reads to build a contig, the similarity between
repeat-region reads can result in the assembly collapsing into an artificially
short sequence or being ignored by more conservative assembly algorithms. Such
regions can be resolved by using algorithms \add{which} specifically reassemble
the collapsed repeat region correctly \cite{mulyukov2002,koren2010}, while
a related approach uses unassembled sequence reads matching the regions around
a scaffold gap to construct a uniquely overlapping set of reads across it
\cite{tsai2010}. \pb
